Estimation of all-cause excess mortality by age-specific mortality patterns of COVID-19 pandemic in Peru in 2020

### 2021-03-23 article.Rmd
primer upload

---
title: Estimation of all-cause excess mortality by age-specific mortality patterns of COVID-19 pandemic in Peru in 2020
author:
- Peter Lloyd-Sherlock^[University of East Anglia]
- Ramón Martínez^[Pan American Health Organisation]
- Shah Ebrahim^[London School of Hygiene and Tropical Medicine]
- Martin McKee^[London School of Hygiene and Tropical Medicine]
- Enrique Acosta ^[Max Plank Institute for Demographich Research]
- Lucas Sempé^[University of East Anglia - email:l.sempe@uea.ac.uk]
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  bookdown::pdf_document2: 
    latex_engine: pdflatex
    toc: no
    fig_caption: yes
    keep_tex: yes
    number_sections: true
  redoc::redoc: default
  bookdown::word_document2: default
  bookdown::html_document2: default
bibliography: Excess.deaths.bib
csl: vancouver.csl
includes:
      in_header: "preambles.tex"
---

@@

We estimate that registered mortality nationally underestimates mortality by `r pm`% (95% CI `r pml`% - `r pmu`%). We estimate total all-cause excess mortality during the period of analysis at `r et` (95% CI  `r etl` - `r etu`) of which `r er` (95% CI  `r erl` - `r eru`) were captured by the vital registration system.  Deaths at age 60 and over accounted for `r age60t`% (95% CI `r age60l`% - `r age60u`%) of total excess deaths, while there were fewer deaths than expected in younger age groups. Lima region, on the Pacific coast and including the national capital, accounts for `r limat` (95% CI  `r limal` - `r limau`) excess deaths, while the regions of Apurimac and Pasco, in the Andes, account for less than 400 excess deaths.

@@

\begin{equation}
  (\#eq:registered)
  \begin{aligned}
  log(\text{Deaths}_{i})=\beta_0+\beta_1\text{COVID-19}+\beta_2\text{Year}+
  \beta_3\epsilon_{t-1}+
  \sum\beta_k\phi_{k}(\text{Week})+
  \text{log}(\text{Pop})+\epsilon_{t}
   \end{aligned}  
\end{equation}

where we fit a natural cubic B-spline function $\phi_{k}(\text{Week})$ on weeks to address long-term trends and seasonality [@Currie2004] and $\text{Year}$ to deal with registration growth. Additionally, we use lagged residuals $\epsilon_{t-1}$ and the log of population in 2020 as an offset. Finally, we compute a dichotomous variable starting in the week corresponding to March 16 2020 as the first registered case of COVID-19 in the country.

@@

\begin{equation}
(\#eq:rr)
\begin{aligned}
\widehat{{\text{Excess}}_\text{Reg}} = (\text{RR}–1)/\text{RR}*\text{n}
\end{aligned}
\end{equation}

@@

exineidep.q<-tateti6.t%>%group_by(Departamento) %>%   
   summarise(`Total excess (TE)`=sum(excess.total.mean,na.rm = T),
            `TE - Lower 95% CI`=sum(excess.total.low,na.rm = T),
            `TE - Upper 95% CI`=sum(excess.total.up,na.rm = T),
     `Excess registered (ER)`=sum(excess_deaths.sum),
     `ER - Lower 95% CI`=sum(excess.low),
             `ER - Upper 95% CI`=sum(excess.up),
     `Excess Covid-19`=sum(complete.covid)) %>%
    adorn_totals()
#knitr::kable(exineidep.q[,1:7],booktabs = T, format="latex",            format.args = list(big.mark = ","),             caption = "Estimated of total excess of deaths by regions",             col.names = c("Region" , "Total excess" , "Lower.CI95%" , "Upper.CI95%" , "Excess reg" , "Covid reg" , "Unregistered")) %>%        kableExtra::kable_styling(latex_options = "hold_position",   font_size = 9 )
colnames(exineidep.q)[1]<-"Region"

@@

exineiageq<-tateti6.t%>%group_by(range) %>%   
  summarise(`Total excess (TE)`=sum(excess.total.mean,na.rm = T),
            `TE - Lower 95% CI`=sum(excess.total.low,na.rm = T),
            `TE - Upper 95% CI`=sum(excess.total.up,na.rm = T),
     `Excess registered (ER)`=sum(excess_deaths.sum),
     `ER - Lower 95% CI`=sum(excess.low),
             `ER - Upper 95% CI`=sum(excess.up),
     `Excess Covid-19`=sum(complete.covid)) %>%
    adorn_totals()

@@

Figure \@ref(fig:map2) shows all cause age-standardised death rates.  We find that `r adj.rates.mores` (CI 95% ranging from `r adj.rates.ci.min` to `r adj.rates.ci.max`) present the highest rates, ranging from `r min` to `r max`, while regions of `r adj.rates.less` show the lowest rates ranging from `r min2` to `r max2` (CI 95% ranging from `r adj.rates.ci.l.min` to `r adj.rates.ci.l.max`) (left). We find that `r dif` shows `r dif.n` deaths per 1000 in comparison to INEI's projections for 2020, which suggests that INEI overestimates mortality rates in Amazonas (right.

@@

>>>>

---
title: Estimation of all-cause excess mortality by age-specific mortality patterns of COVID-19 pandemic in Peru in 2020
author:
- Peter Lloyd-Sherlock^[University of East Anglia]
- Ramón Martínez^[Pan American Health Organisation]
- Shah Ebrahim^[London School of Hygiene and Tropical Medicine]
- Martin McKee^[London School of Hygiene and Tropical Medicine]
- Enrique Acosta ^[Max Planck Institute for Demographich Research]
- Lucas Sempé^[University of East Anglia - email:l.sempe@uea.ac.uk]
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  bookdown::pdf_document2: 
    latex_engine: pdflatex
    toc: no
    fig_caption: yes
    keep_tex: yes
    number_sections: true
  bookdown::word_document2: default
  bookdown::html_document2: default
bibliography: Excess.deaths.bib
csl: vancouver.csl
includes:
      in_header: "preambles.tex"
---

@@

We estimate that registered mortality nationally underestimates mortality by `r pm`% (95% CI `r pml`% - `r pmu`%). We estimate total all-cause excess mortality during the period of analysis at `r et` (95% CI  `r etl` - `r etu`) of which `r er` (95% CI  `r erl` - `r eru`) were captured by the vital registration system.  Deaths at age 60 and over accounted for `r age60t`% (95% CI `r age60l`% - `r age60u`%) of total excess deaths, while there were fewer deaths than expected in younger age groups. Lima region, on the Pacific coast and including the national capital, accounts for `r limat` (95% CI  `r limal` - `r limau`) excess deaths, while the regions of `r menos5` account for less than `r menos5.num` excess deaths.

@@

\begin{equation}
  (\#eq:registered)
  \begin{aligned}
  log(\text{Deaths}_{i})=\beta_0+\beta_1\text{COVID-19}+\beta_2\text{Week}+\beta_3\text{Year}+
  \beta_4\epsilon_{t-1} + \text{log}(\text{Pop})+\epsilon_{t}
   \end{aligned}  
\end{equation}

where we fit weekly counts of deaths $\text{Week}$ and a variable $\text{Year}$ to address long-term trends. Additionally, we use lagged residuals $\epsilon_{t-1}$ and the log of population in 2020 as an offset. Finally, we compute a dichotomous variable starting in the week corresponding to March 16 2020 as the first registered case of COVID-19 in the country.

@@

\begin{equation}
(\#eq:rr)
\begin{aligned}
\widehat{{\text{Excess}}_\text{Reg}} = (\text{RR}-1)/\text{RR}*\text{n}
\end{aligned}
\end{equation}

@@

exineidep.q<-tateti6.t%>%group_by(Departamento) %>%   
   summarise(`Total excess (TE)`=sum(excess.total.mean,na.rm = T),
            `TE - Lower 95% CI`=sum(excess.total.low,na.rm = T),
            `TE - Upper 95% CI`=sum(excess.total.up,na.rm = T),
     `Excess registered (ER)`=sum(excess_deaths.sum),
     `ER - Lower 95% CI`=sum(excess.low),
             `ER - Upper 95% CI`=sum(excess.up),
     `Excess Covid-19`=round(sum(complete.covid)),0) %>%
    adorn_totals()
#knitr::kable(exineidep.q[,1:7],booktabs = T, format="latex",            format.args = list(big.mark = ","),             caption = "Estimated of total excess of deaths by regions",             col.names = c("Region" , "Total excess" , "Lower.CI95%" , "Upper.CI95%" , "Excess reg" , "Covid reg" , "Unregistered")) %>%        kableExtra::kable_styling(latex_options = "hold_position",   font_size = 9 )
colnames(exineidep.q)[1]<-"Region"

@@

exineiageq<-tateti6.t%>%group_by(range) %>%   
  summarise(`Total excess (TE)`=sum(excess.total.mean,na.rm = T),
            `TE - Lower 95% CI`=sum(excess.total.low,na.rm = T),
            `TE - Upper 95% CI`=sum(excess.total.up,na.rm = T),
     `Excess registered (ER)`=sum(excess_deaths.sum),
     `ER - Lower 95% CI`=sum(excess.low),
             `ER - Upper 95% CI`=sum(excess.up),
     `Excess Covid-19`=round(sum(complete.covid),0)) %>%
    adorn_totals()

@@

Figure \@ref(fig:map2) shows all cause age-standardised death rates.  We find that `r adj.rates.mores` (CI 95% ranging from `r adj.rates.ci.min` to `r adj.rates.ci.max`) present the highest rates, ranging from `r min` to `r max`, while regions of `r adj.rates.less` show the lowest rates ranging from `r min2` to `r max2` (CI 95% ranging from `r adj.rates.ci.l.min` to `r adj.rates.ci.l.max`) (left). We find that `r dif` shows `r dif.n` deaths per 1000 in comparison to INEI's projections for 2020, which suggests that INEI overestimates mortality rates in Amazonas (right).

@@

# Contributors

LS conceived and initiated the study, did the statistical analysis and visualisations and drafted the manuscript.

PLS conceived and supervised the study, drafted the manuscript and led the editing process of the manuscript.

MM and SE supervised the study and engaged in the draft review & editing

RM contributed in the data curation and methodology process and reviewed & edited the draft.

EA contributed in data curation, methodology and statistical analysis.

LS and EA had access to all the data. All authors approved the manuscript and are responsible for the decision to submit for publication.

# Declaration of interests

All other authors declare no competing interests.

# Data sharing

Data presented in this manuscript are made publicly available at https://github.com/lsempe77/excess

# Acknowledgments

Any?
